Smoller Brainspan Analysis
========================================================


```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, echo=TRUE, highlight=TRUE, figalign='center', fig.height=9, fig.width=9, out.width='800px', message=FALSE, error=TRUE, warning=FALSE, cache=FALSE)

# Setup report details
clientname="Erin Dunn"
clientemail="erindunn@pngu.mgh.harvard.edu"
lablocation="MGH"
analystname="Meeta Mistry"
analystemail="mmistry@hsph.harvard.edu"
```

Array analysis for `r clientname` (`r clientemail`) at `r lablocation`. Contact `r analystname` (`r analystemail`) for additional details. Request from client was:

> It doesn't look like BrainCloud is an ideal dataset because of the issues related to the PMI. Given this, I think we should probably focus on BrainSpan and run a parallel set of analyses to see if things look cleaner in that new dataset. 

## Workflow: 
* grab the BrainSpan data set and metadata
* re-run the QC of the metadata and basic expression analysis 
* run STEM on one brain region only

## Setup

### Bioconductor and R libraries used

```{r libraries, echo=TRUE}
library(ggplot2)
library(gtable)
library(scales)
library(RColorBrewer)
library(GEOquery)
library(affy)
library(arrayQualityMetrics)
library(reshape)
library(xtable)
library(ruv)
library(limma)
library(Biobase)
library(gridExtra)

```

### Get variables
- get base directory for analyses
- specify data and results directories
- specify column headers used in metadata file


```{r variables, echo=TRUE}
# Setup directory variables
baseDir <- '.'
dataDir <- file.path(baseDir, "data")
metaDir <- file.path(dataDir, "meta")
resultsDir <- file.path(baseDir, "results")
```


### Load the expression data

```{r dataimport GEO, echo=TRUE}

# Load GEO data

# Affymetrix GeneChip Mouse Genome 430 2.0 microarrays. Array data was processed using the Affy and gcrma packages
gse1 <- getGEO('GSE9202', destdir=file.path(dataDir, 'geo'))
```


